---
title: "Covariates Comparison"
author: "Michael Karcher"
date: "May 20, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple BNPR Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette shows a simple and typical use of `phylodyn` 
and its main functions BNPR and BNPR_PS, 
contrasting different sampling time models.
We start by loading the `phylodyn` package and setting a random seed.

```{r}
library(phylodyn)
set.seed(8675311)
```

We set the different trajectories:
`traj1(t)` is the effective population size,
`traj2(t)` is a covariate trajectory,
and the product of those two functions `trajc(t)` is the sampling intensity.

```{r}
beta1 = 1
beta2 = 1

traj1 = function(t) logistic_traj(t, offset = 6) / 5
traj2 = function(t) exp_traj(t = t, scale = 1, rate = 0.05)
trajc = function(t) traj1(t)^beta1 * traj2(t)^beta2
```

We plot each trajectory to illustrate. 
Note: the present (`t=0`) is on the right, and time flows from left to right.

```{r fig.width=16, fig.height=6, out.width='100%'}
ylim=c(0.5,100)
par(mfrow=c(1, 3), oma=c(4,0,0,0)+0.1, mar=c(0.5,2,2,1), 
    xpd=NA, cex=1.2, cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
curve(traj1(t), xlim=c(60, 0), xname = "t", main = "Effective population size (traj1)", ylab = "")
curve(traj2(t), xlim=c(60, 0), xname = "t", main = "Covariate (traj2)", ylab = "")
curve(trajc(t), xlim=c(60, 0), xname = "t", main = "Sampling intensity (trajc)", ylab = "")
```

We simulate sampling times according to the sampling intensity.

```{r fig.width=6, fig.height=6, out.width='50%'}
samp_end = 60
nsamps   = 1000
Cprop      = nsamps/integrate(traj_beta, 0, samp_end, traj=trajc, beta=1)$value

samp_times = pref_sample(trajc, c=Cprop, lim=c(0,samp_end), beta=1)
hist(samp_times, xlim = c(samp_end, 0))
```

We simulate a genealogy using a coalescent model.

```{r}
gene = coalsim(samp_times = samp_times, n_sampled = rep(1, length(samp_times)), traj = traj1)
```

We fit three models and plot the results:

* Sampling conditional (no sampling time model)
* Sampling aware, no covariates
* Sampling aware, one covariate (exponential growth)

```{r warning=FALSE, message=FALSE, fig.width=16, fig.height=6, out.width='100%'}
res_BNPR = BNPR(data = gene, lengthout = 100)
res_BNPR_PS1 = BNPR_PS(data = gene, lengthout = 100)
res_BNPR_PS2 = BNPR_PS(data = gene, lengthout = 100, fns = list(traj2))

ylim=c(0.5,100)
trev_blue = rgb(0.330, 0.484, 0.828)
trev_yell = rgb(0.829, 0.680, 0.306)
trev_purp = rgb(0.533, 0.000, 0.607)

par(mfrow=c(1, 3), oma=c(4,3,0,0)+0.1, mar=c(0.5,2,2,1), 
    xpd=NA, cex=1.2, cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)

plot_BNPR(res_BNPR, traj=traj1, ylim=ylim, lwd=3, 
          col=trev_yell, traj_col = "black", traj_lty = 2, 
          main="BNPR", heatmap_labels = FALSE)
text(x = 12, y = 50, labels = "Estimate", col = trev_yell, cex = 1.0, font = 2)
arrows(x0 = 12, y0 = 40, x1 = 3, y1 = 20, length = 0.1, lwd = 2, col = trev_yell)
text(x = 35, y = 90, labels = "Cred. region", col = "darkgray", cex = 1.0, font = 2)
arrows(x0 = 35, y0 = 75, x1 = 44, y1 = 30, length = 0.1, lwd = 2, col = "darkgray")

plot_BNPR(res_BNPR_PS1, traj=traj1, ylim=ylim, lwd=3, 
          col=trev_blue, traj_col = "black", traj_lty = 2, 
          ylab="", main="BNPR-PS", heatmap_labels = FALSE)
text(x = 12, y = 50, labels = "Estimate", col = trev_blue, cex = 1.0, font = 2)
arrows(x0 = 12, y0 = 40, x1 = 3, y1 = 25, length = 0.1, lwd = 2, col = trev_blue)
text(x = 55, y = 50, labels = "True Trajectory", col = "black", cex = 1.0)
arrows(x0 = 55, y0 = 40, x1 = 59, y1 = 21, length = 0.1, col = "black", lty = 1, lwd=2)

plot_BNPR(res_BNPR_PS2, traj=traj1, ylim=ylim, lwd=3, 
          col=trev_purp, traj_col = "black", traj_lty = 2, 
          ylab="", main="BNPR-PS with Cov", heatmap_labels = TRUE, 
          heatmap_labels_side = "left", heatmap_labels_cex = 1.0)
text(x = 12, y = 50, labels = "Estimate", col = trev_purp, cex = 1.0, font = 2)
arrows(x0 = 12, y0 = 40, x1 = 3, y1 = 22, length = 0.1, lwd = 2,
       col = trev_purp)
```

We note that including a sampling time model without the correct covariates
can result in worse results than not including a sampling time model at all,
but including the right covariate provides better results.
We pursue the question of model adequacy using posterior predictive checks
in Karcher, et al. 2019 and in the vignette `SimplePPC.Rmd` in this package.

## References

1. Karcher MD, Suchard MA, Dudas G, Minin VN (2019). Estimating effective population size changes from preferentially sampled genetic sequences, [arXiv:1903.11797](https://arxiv.org/abs/1903.11797).

2. Palacios JA and Minin VN.
Integrated nested Laplace approximation for Bayesian nonparametric phylodynamics.
In *Proceedings of the Twenty-Eighth International Conference on Uncertainty in Artificial Intelligence*, pages 726–735, 2012.

3. Lan, S, Palacios, JA, Karcher, M, Minin, VN, & Shahbaba, B (2014).
An Efficient Bayesian Inference Framework for Coalescent-Based Nonparametric Phylodynamics.
 *Bioinformatics*, 31, 3282-3289.
