---
title: "Bayesian Nonparametric Inference of Population Size Changes from Sequential Genealogies"
author: "Julia Palacios"
date: "October 12, 2015"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phylodyn from local genealogies}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Bayesian Nonparametric Inference of Population Size Changes from Sequential Genealogies
==========================================================================


This R code implements the method described in "Bayesian Nonparametric Inference of Population Size Changes from Sequential Genealogies" by Palacios JA, Wakeley J, and Ramachandran S (doi: http://dx.doi.org/10.1101/019216). The method is applied to a test dataset (data(Bottle_20c.txt)).


**Newick files preparation (optional)**
----------------------------

This program takes a file of local genealogies (with n>2) in [Newick](http://evolution.genetics.washington.edu/phylip/newicktree.html) format as input. Our example dataset  is in the data folder (Bottle_20c.txt). In this example, we use [MaCS](https://code.google.com/p/macs/) with the following command lines

```
./macs 20 300000 -t 4.0 -eN 0 1 -eN 0.3 0.1 -eN 0.5 1 -T -r .002 -h 1 -s 1420826310 >outMacs.tree
```
The file outMacs.tree created above contains more information than needed for our analysis, which only requires gene genealogies. With the following command, we extract the gene genealogies in Newick format
```
awk '$1~"NEWICK_TREE"' outMacs.tree | awk -F\] '{print $2}' >Bottle_20.txt
```

**R Code**
---------------
We start by loading the library and reading our data
```{r}
library("phylodyn")
data("Bottle_20c")
```
Our method assumes that genealogies are a realization of the Sequentially Markov Coalescent model [SMC'](http://bmcgenet.biomedcentral.com/articles/10.1186/1471-2156-7-16)



**Data Preparation**
----------------------

For this example, we will run our algorithm on the first 50 local genealogies (sim=50) and scale time by 10 (scaling=10). Our algorithm searches the new and deleted coalescent times by comparing the coalescent times of consecutive genealogies and we define a tolerance (tol=.00001) to set whether two coalescent times are different ($t_{a} \neq t_{b}$ if $|t_{a}-t_{b}|>tol$). In our experiments, a tolerance level of .00001 works well when the time to the most recent common ancestor is of the order of 2-10. 

```{r}
sim<-length(Bottle_20c) 
sim
sim<-50 #For this example, the first 50 genealogies
scaling<-10
tol<-.00001 #tolerance factor to detect difference between branch lengths
D<-read_times(Bottle_20c,sim,scaling)
```

Matrix D is a matrix with sim=50 rows and n-1=19 columns with coalescent times. To see the summary of the time to the most recent common ancestor time, run the command:

```{r}
summary(D[,dim(D)[2]])
```

Next, we define our discretization of the population size function. After testing many different number of change points, we find that 100 regularly spaced change points provides a good resolution of $N(t)$

```{r}
window<-max(D)+.0001
grid.size<-100
grid<-seq(0,window,length.out=grid.size)
grid<-c(grid,max(D)+.0002)
```

We then adjust all our sufficient statistics for our chosen discretization  

```{r}
info<-find_info2(Bottle_20c,D,sim,tol,scaling)
```

**MCMC Sampling**
-----------------------

For all our results in the manuscript, we used the seed value 2014 with 50000 iterations (NSAMP=50000) and a burnin of 1000 iterations (NBURNIN=1000). For this tutorial we use 50 iterations and 5 iterations of burnin. 

```{r}
set.seed(2014)
res_MCMC = smcp_sampling(data = info, nsamp = 50,
                         nburnin = 5,grid)

```

**Summary of Results**
--------------------------

We compute the posterior median and 95% BCIs of log N(t) and plot the results:

```{r}

##Plot results
results<-res_MCMC
plot(results[,1],results[,3],type="l",xlim=c(1,0),ylim=c(-3,3),ylab="log N(t)",xlab="No generations",col="white")
plot_res(results) 

##True trajectory
x<-sort(c(0.299999,0.3,0.49999,0.5,seq(0,4,length.out=100)))
y<-x
y[x<.3]<-log(.5)
y[x>=.3 & x<.5]<-log(.1/2)
y[x>=.5]<-log(.5)
points(x,y,lty=2,type="l",lwd=1.5)

```  

References
---
1. Palacios, JA, Wakeley, J, and Ramachandran, S. Bayesian nonparametric inference of population size changes from sequential genealogies. *Genetics* 2015 Vol. 201:281-304


