---
title: "PC prior"
author: "Luiz Max Carvalho"
date: "July 25, 2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PC prior}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
This vignette illustrates the employment of a penalised complexity (PC) prior (Simpson et al., 2017) on the GMRF precision, $\kappa$.
The precision controls the smoothness of the population size changes: low precision (high variance) means abrupt changes between coalescent events and high precision (low variance) means a gradual change in population sizes.
Statistically, we want the _right_ level of smoothness: too little and we are fitting noise; too much and we lose important temporal variation.

It is thus important to place a well-thought-out prior on the precision.
The main idea is to elicit ("create") a prior distribution that controls the standard deviation of the (log) population sizes without leading to overfitting (see Simpson et al., 2017, Section 3.3).
It can be shown that the usual Gamma prior leads to overfitting.
In particular, the default Gamma prior has shape and rate parameters equal to $0.001$, leading to a flat prior distribution on $\kappa$.

To this end, the authors show that a prior that penalises complex models and avoids overfitting is the Gumbel type II distribution.
Let $a, b > 0$ be the shape and scale hyperparameters respectively; the probability density function is then
$$
 \pi_2(\kappa \mid a, b) = ab \cdot \kappa^{-a-1}\exp\left(-b\kappa^{-a}\right),\: \kappa > 0.
$$
The authors recommend $a = 1/2$ and $b$ to be chosen such that $\textrm{Pr}(1/\sqrt{\kappa} > S) = p$, where the value $S$ and the probability $p$ are to be chosen on substantive grounds.
This leads to $b = -\ln(p)/S$.
The defaults in `BNPR()` when `pc_prior=TRUE` are $S=1$ and $p = 0.1$. 

First, let us set up the hyperparameters for the PC prior:
```{r hyperpar}
S <- 1
p <- .1
apc <- 1/2
bpc <- -log(p)/S
```

To make things more interesting (and fair), we will also elicit a Gamma distribution with the constraint that $\textrm{Pr}(1/\sqrt{\kappa} > 1) = 0.1$. 
This leads to a Gamma distribution with $\textrm{shape}=2.544882$ and $\textrm{scale}=1.203437$, which we will call a "matching" Gamma.

Now, a comparison between the priors:
```{r compfig, fig.width=8, fig.height=6,echo=FALSE}
dgumbel2 <- function(x, a, b, log = FALSE){
  ans <- log(a) + log(b) -( a + 1)*x -b*x^-a
  if(!log) ans <- exp(ans)
  return(ans)
}
dgumbel2 <- Vectorize(dgumbel2)

curve(dgamma(x, shape = 2.544882, scale = 1.203437),
      0, 10,
      lwd = 4, lty = 2,
       xlab = expression(kappa), ylab = "Density")
curve(dgumbel2(x, a = apc, b = bpc), 0, 10,
       lwd = 4, lty = 3, add = TRUE)
curve(dgamma(x, shape = 1E-3, rate = 1E-3),
      lwd = 4, add = TRUE)
legend(x = "topright",
       cex = 1.5,
       legend = c("Default Gamma",
                  "Matching Gamma",
                  "Gumbel type II (PC)"),
       lwd = 3, lty = 1:3, bty = 'n')
```

Now we will see these priors in action in the analysis of the New York flu data:
```{r analysis}
library(phylodyn)

data("NY_flu")

res.default <- BNPR(NY_flu)
res.RG <- BNPR(NY_flu,
               prec_alpha =  2.544882, prec_beta = 1/1.203437)
res.PC <-  BNPR(NY_flu, pc_prior = TRUE)
```

Extracting the reconstructions and plotting:
```{r reconstructions, fig.width=8, fig.height=6}
all.reconstructions <- do.call(rbind,
                               list(
                                 data.frame(res.default$summary,
                                            prior = "default_Gamma"),
                                 data.frame(res.RG$summary,
                                            prior = "matching_Gamma"),
                                 data.frame(res.PC$summary,
                                            prior = "penalised_complexity")
                               ))

library(ggplot2)

ggplot(data = all.reconstructions,
            aes(x = time, y = mean,
                colour = prior, fill = prior)) +
  geom_line(size = 1) +
  scale_y_log10("Effective population size") +
  scale_x_continuous("Time") + 
  geom_ribbon(aes(ymin = quant0.025,
                  ymax = quant0.975),
              alpha = 0.4) + 
  theme_bw(base_size = 20)
```
As we can see, the matching Gamma prior leads to a posterior with way more uncertainty. 
Slightly harder to see is that the posterior under the PC prior is the one with the lowest uncertainty.

## References

1. Simpson, D., Rue, H., Riebler, A., Martins, T. G., & Sørbye, S. H. (2017). Penalising model component complexity: A principled, practical approach to constructing priors. Statistical Science, 32(1), 1-28.