---
title: "Simple Posterior Predictive Check Vignette"
author: "Michael Karcher"
date: "May 15, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple Posterior Predictive Check Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette shows a simple and typical use of posterior predictive checks in `phylodyn` 

We start by loading the `phylodyn` package.

```{r message=FALSE}
set.seed(8675309)
library(phylodyn)
```

We need to set the true effective population size trajectory function. For this example, we choose a seasonal growth/collapse trajectory (already implemented in `phylodyn`).

```{r}
traj = function(t)
{
  return(100*exp(-0.5*t))
}
```

We set the end of the sampling interval and the number of samples we should expect.

```{r}
samp_end = 5
nsamps   = 100
```

We calculate the proportionality constant necessary to produce the correct expected number of samples. We then produce a set of samples by drawing from an inhomogeneous Poisson process with intensity proportional to the effective population size trajectory.

```{r}
Cprop      = nsamps/integrate(traj_beta, 0, samp_end, traj=traj, beta=1)$value
samp_times = pref_sample(traj, c=Cprop, lim=c(0,samp_end), beta=1)
```

We simulate a genealogy based on our sample using the coalescent.

```{r}
gene = coalsim(samp_times = samp_times, 
               n_sampled = rep(1, length(samp_times)),
               traj = traj)
```

We set the number of samples and burn-in parameters. For expediency in this vignette we set them to be small. To produce more meaningful results, increase both parameters.

```{r}
nsamp = 2000
nburnin = 1000
```

We invoke the `mcmc_sampling` function with an elliptical slice sampler for the population parameters and with a Metropolis Hastings sampler for the sampling model parameters.

```{r results='hide'}
res_MCMC = mcmc_sampling(data = gene, alg = 'ESS', nsamp = nsamp,
                         nburnin = nburnin, samp_alg = "MH")
```

We plot the results from the MCMC.

```{r fig.width=6, fig.height=6, out.width='50%'}
plot_MCMC(res_MCMC, traj = traj, main = "MCMC Output")
legend('topleft',c('Truth', 'MCMC'), bty='n', lty=c(2,1))
```

We note that the MCMC results include the log-effective population sizes as well as any hyperparameters,
so we separate and store a matrix of the log-effective population sizes.

```{r}
colnames(res_MCMC$samp)
logpop_mat = res_MCMC$samp[,1:99]
betas_mat =  res_MCMC$samp[,101:102]
```

We run the coalescent posterior predictive check here and calculate the posterior predictive p-value.
A value close to 0 or 1 would suggests model inadequacy.

```{r}
coal_ppc = posterior_coal_check(coal_times = res_MCMC$coal_times, 
                                samp_times = res_MCMC$samp_times,
                                n_sampled = res_MCMC$n_sampled, 
                                logpop = logpop_mat, 
                                grid = res_MCMC$grid)
mean(coal_ppc$rep > coal_ppc$obs)
```

We plot the results, and note how the $y=x$ line goes through the cloud of points.

```{r fig.width=6, fig.height=6, out.width='50%'}
plot(coal_ppc$obs, coal_ppc$rep, pch = 16, col = rgb(0, 0, 0, 0.5),
     main = "Coalescent PPC", xlab = "", ylab = "")
title(ylab = "Replicated", line = 2)
title(xlab = "Observed", line = 2)
abline(a = 0, b = 1)
```

We repeat the analysis for the sampling time posterior predictive check.

```{r}
samp_ppc = posterior_samp_check(samp_times = res_MCMC$samp_times,
                                n_sampled = res_MCMC$n_sampled, 
                                logpop = logpop_mat, 
                                grid = res_MCMC$grid,
                                betas = betas_mat)
mean(samp_ppc$rep > samp_ppc$obs)
```

We plot the results, and note how the $y=x$ line goes through the cloud of points.

```{r fig.width=6, fig.height=6, out.width='50%'}
plot(samp_ppc$obs, samp_ppc$rep, pch = 16, col = rgb(0, 0, 0, 0.5),
     main = "Sampling-aware, Sampling PPC", xlab = "", ylab = "")
title(ylab = "Replicated", line = 2)
title(xlab = "Observed", line = 2)
abline(a = 0, b = 1)
```

## References

1. Karcher MD, Suchard MA, Dudas G, Minin VN (2019). 
Estimating effective population size changes from preferentially sampled genetic sequences. 
[arXiv:1903.11797](https://arxiv.org/abs/1903.11797).

2. Palacios JA and Minin VN (2012).
Integrated nested Laplace approximation for Bayesian nonparametric phylodynamics.
In *Proceedings of the Twenty-Eighth International Conference on Uncertainty in Artificial Intelligence*, 726–735.

3. Lan, S, Palacios, JA, Karcher, M, Minin, VN, & Shahbaba, B (2014).
An Efficient Bayesian Inference Framework for Coalescent-Based Nonparametric Phylodynamics.
 *Bioinformatics*, 31, 3282-3289.
