---
title: "Coalescent simulation of genealogies"
author: "Julia Palacios"
date: "October 6, 2016"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
Coalescent simulation of genealogies
========================================================


A Genealogy (tree) consist of Topology and coalescent times. 

* Topology can be *isochronous* (all tips are "sampled" at time 0) or *heterochronous* (tips have different "sampling times"). Lineages merge (coalesce) at random between the existant lineages at the coalescent times. 
* Coalescent times are exponentially distributed random variables with a rate that depends on the number of lineages and the population size trajectory.


The main function for simulation of genealogies in phylodyn is \texttt{coalsim}. In all our simulations, time is measured in units of $N_{0}$ generations.

gene = coalsim(samp_times = samp_times, n_sampled = n_sampled, traj = traj, lower_bound = 1/20)

Examples
==================================
The following example generates an isochronous genealogy from a constant effective population size (Ne=1) with 10 tips.
```{r}

library("phylodyn")

constant<-function(x){
  return (rep(1,length(x)))
}

simulation1<-coalsim(0,10,constant)
tree1<-generate_newick(simulation1)
plot(tree1$newick)
tree1$newick
```

In the following example we simulate a genealogy with constant effective population size with 10 tips at time 0 and 40 other tips with uniform sampling times 
```{r}

samp_times = c(0, sort(runif(40, 0, 8)))
n_sampled = c(10, rep(1, 40))

simulation2<-coalsim(samp_times = samp_times, n_sampled = n_sampled, traj = constant)
tree2<-generate_newick(simulation2)
plot(tree2$newick,show.tip.label = F)
tree2$newick

```

Here, we consider a more general demographic model such as bottleneck. We specify our trajectory through the function \texttt{bottleneck_traj} and simulate via the thinning method (see Palacios and Minin, 2013)
```{r}

bottleneck_traj<-function(t){
  result=rep(0,length(t))
  result[t<=0.5]<-1
  result[t>0.5 & t<1]<-.1
  result[t>=1]<-1
  return(result)
}

simulation3<-coalsim(samp_times = samp_times, n_sampled = n_sampled, traj = bottleneck_traj,method="thin",val_upper=11)
tree3<-generate_newick((simulation3))
plot(tree3$newick,show.tip.label = F)
```

Note: Other R packages such as Ape and phyclust (ms) have functions that simulate isochronous genealogies under the coalescent model for specific demographic scenarios such as exponential growth and piece-wise constant functions. 





